---
title: "final project"
author: "Fanding Zhou"
date: "4/29/2022"
output: pdf_document
---
\newcommand{\indep}{\perp \!\!\! \perp}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggdag)
library(dagitty)
```

## Step 1: State scientific question.


(a) The causal question

What is the effect of a large-scale primary care redesign, the Comprehensive Primary Care Plus (CPC+) Initiative, on the monthly Medicare patient expenditures?


(b) The target population

Patients in different primary care practices in 18 regions. 

## Step 2: Specify a structural causal model (SCM). 

(a) The endogenous variables V are variables that are meaningful for the scientific question and affected by other variables in this model. For this study, we define the following endogenous variables:

* $L_0 = \{V1, V2, V3, V4, V5\}_{t=2}$

* $L_1 = \{V1, V2, V3, V4, V5\}_{t=3}$

* V1: age, continuous variable supported on [0,100]

* V2: income groups, ordinal variable encoded into 0~14 integers.

* V3: sex, 0 for female, 1 for male, appromately equally distributed.

* V4: mean standardized Hierarchical Condition Category (HCC) score, approximately standard normal distribution.

* V5: race categories, white/black/all other, 70:20:10.

* V1-V5 are time variant patient-level covariates

* $X = \{X1, X2, X3, X4, X5, X6, X7, X8,X9\}$

* X1-X9 are time-invariant practice-level covariates.

* $Z$ = Initiative of Comprehensive Primary Care Plus (CPC+) in year 3


* $Y_0$ = Monthly Medicare patient expenditures in year 2

* $Y_1$ = Monthly Medicare patient expenditures in year 3

Here $V = (C_0,C_1,T_0,T_1,Y_0,Y_1)$


(b) The exogenous U include all the unmeasured or unknown factors not included in V that impact the values that the V variables take. 

The exogenous nodes are $U = (U_X,U_T,U_{L_0},U_{L_1},U_{Y_0},U_{Y_1})$. 

According to the guidance of ACIC competition, the DGPs will be free of unmeasured confounding. In other words, we can place some independence assumptions on the distribution of unmeasured factors $P_U$. To simplify, we assume the all the exogenous variables satisfy pairwise independence.


(c)  This would suggest the following structural equations F:

$$X =f_{X}(U_{X})$$
$$L_0 =f_{L_0}(U_{L_0})$$
$$Y_0 = f_{Y_0}(U_{Y_0},L_0)$$
$$L_1 =f_{L_1}(U_{L_1},L_0,Y_0)$$
$$Z = f_Z(U_{Z},L_1,Y_0,X)$$
$$Y_1 = f_{Y_1}(U_{Y_1},L_1,T,Y_0,X)$$
(d) Exclusion restrictions. 

An exclusion restriction is when a variable is not directly affected by another variable that precedes it in the structural causal model. For our model, we assume $L_0$ didn't directly affect $T$, but has some indirect effect through $L_1$ and $Y_0$. In addition, we assume practice level covariants X only have impact on $Y_0$, $Y_1$, $T$, but do not on those time-variant patient level covariants $L_0$ and $L_1$.

(f) Causal graph.

```{r}
dag_m <- dagitty('dag {
    Y0 [pos="0,0"]
    L0 [pos="0,1"]
    Z [pos="0.5,0.5"]
    X [pos="0.5,0"]
    L1 [pos="0.5, 1"]
    Y1 [pos="1,0.5"]
    UL0[pos="-0.3,1"]
    UL1[pos="0.5,1.5"]
    UY0[pos="-0.3,0"]
    UZ[pos="0.75,1"]
    UX[pos="0.75,0"]
    UY1[pos="1,1"]
    UL0->L0
    UL1->L1
    UY0->Y0
    UZ->Z
    UX->X
    UY1->Y1
    L0->L1
    L0->Y0
    Y0->Z
    Y0->L1
    Y0->Y1
    L1->Y1
    L1->Z
    X->Y0
    X->Z
    X->Y1
    Z->Y1
    }')
ggdag(dag_m, layout = "circle")

```



## Step 3. Specify the target parameter of the observed data distribution 

(a) The counterfactuals of interest are Y(1) and Y(0) and they can be defined as 

- $Y_1(1)$: Monthly Medicare patient expenditures if all patients had joined CPC+ in year 3.

- $Y_1(0)$: Monthly Medicare patient expenditures if all patients hadn't joined CPC+ in year 3.

(b) Our target parameter is the average causal effect: 
$$\Psi_{U,V}(P_0) = \mathbb E_{U,V}[Y_1(1) - Y_1(0)]$$ 
This means the target causal parameter is the difference in the monthly Medicare patient expenditures for patients enrolled in CPC+ for 2 years and not enrolled in CPC+.

## Step 4. Specify the observed data.

Suppose the data is from i.i.d sample from n = 329250 randomly sampled
patients.

* $Ti$: whether patient i has enrolled in CPC+ in year 3

* $Xi$: practical level confounders collected for patient i 

* $L_{0,i}$: patient level confounders collected for patient i in year 2
* $L_{1,i}$: patient level confounders collected for patient i in year 3
* $Y_{0,i}$: Monthly Medicare patient expenditures for patient i in year 2
* $Y_{1,i}$: Monthly Medicare patient expenditures for patient i in year 3

Further assume that the collected data are independent and identically distributed. 

## Step 5. Identify the targeted causal effect with the observed data. 
(a) Conditional independent assumption

$$Y_{1,i}(0),Y_{0,i}(0) \indep Z|X,L_1,Y_0$$

(b) By backdoor-path criterion, we can write our causal parameter with the observed data by G-computation: 

$$
\Psi(P_0) = \mathbb E[\mathbb E[Y_1 |T = 1,X,L_1,Y_0]] - \mathbb E[\mathbb E[Y_1 |T = 0,X,L_1,Y_0]]
$$
Under our causal model and assumptions, average treatment effect equals to the observed difference in
mean outcome within confounder strata, standardized to distribution of confounders.

(c) Relevant positivity assumption


## Step 6. Estimation and Statistical Inference.
```{r eval=FALSE, include=FALSE}
library(dplyr)
i = 1

practice_files = list.files("practice", pattern="*.csv", full.names=TRUE)
practice_year_files = list.files("practice_year", pattern="*.csv", full.names=TRUE)
patient_files = list.files("patient", pattern="*.csv", full.names=TRUE)
patient_year_files = list.files("patient_year", pattern="*.csv", full.names=TRUE)
practice = read.csv(practice_files[i])
practice_year = read.csv(practice_year_files[i])
patient = read.csv(patient_files[i])
patient_year = read.csv(patient_year_files[i])
practice_year_exposure = practice_year[, c("id.practice","year","Z")]
mergedata = patient %>%
  dplyr::full_join(patient_year, by = "id.patient") %>%
  dplyr::full_join(practice, by = "id.practice") %>%
  dplyr::full_join(practice_year_exposure, by = c("id.practice","year")) 
  
meta1 = mergedata %>% filter(year == 3)
meta2 = mergedata %>% 
  filter(year == 2) %>%
  select(id.patient,Y)
colnames(meta2)[2]= "Y0"
meta3 = meta1 %>%
  dplyr::inner_join(meta2, by = "id.patient") 
ObsData = data.frame(meta3[,c(paste0("X",1:9),paste0("V",1:5),"Y0")],Z=meta3$Z,Y=meta3$Y)
write.table(ObsData, "filtered_patient_1.csv")

```
```{r}
set.seed(252)
ObsData = read.table("filtered_patient_1.csv")
```


```{r eval=FALSE, include=FALSE}
#------------------------------------------
# 1. Load the Super Learner & ltmle packages
#------------------------------------------
library("SuperLearner")
library("ltmle")
SL.library<- c("SL.glm", "SL.gam") #"SL.step"

#------------------------------------------
# 2. call ltmle with Super Learner (same libraries)
#------------------------------------------
ltmle.SL<- ltmle(data=ObsData, Anodes='Z', Ynodes='Y', abar=list(1,0),
                 SL.library=SL.library)
summary(ltmle.SL)
saveRDS(ltmle.SL,"result1.rds")
```

result: 
```{r}
ltmle.SL = readRDS("result1.rds")
summary(ltmle.SL)
```
